---
title: "Abdussamad - 190519205 - Question 3 - R"
author: "Abdussamad Roshan"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 3 - How does the number of people flying between different locations change over time?

## Importing and Loading Packages

```{r}
# # Installing the necessary packages
install.packages("dplyr")
install.packages("ggplot2")
install.packages("plotly")
install.packages("rmarkdown")
install.packages("scales")
install.packages("tidyverse")
```

```{r}
# Loading the packages
library(dplyr)
library(ggplot2)
library(plotly)
library(rmarkdown)
library(scales)
library(tidyr)
library(tidyverse)
```

## Importing the Datasets

```{r}
# Importing the main cleaned dataset
cleaned_df <- read.csv("Cleaned_Dataset_R.csv")
# Importing the Plane Data, any NA or empty values are replaced with NA
airports_df <- read.csv("airports.csv", na.strings = c("", "NA"))
```

```{r}
# Checking the first 5 rows of the airports dataset
head(airports_df, n = 5)
```

```{r}
# Checking for any duplicated rows in the Airports Dataset
head(airports_df[duplicated(airports_df),])
```

```{r}
# Checking for null values in the airports dataset
colSums(is.na(airports_df))
```

```{r}
# Removing all rows with null values in the airports dataset and checking if they have been removed
airports_df <- na.omit(airports_df)
colSums(is.na(airports_df))
```

States will be used to compare how flights change between locations

```{r}
# Filtering to keep only the neccessary columns in the main dataset
cleaned_df <- cleaned_df[, c("Year", "Origin", "Dest")]
cleaned_df
```

```{r}
# Filtering to keep only the neccessary columns in the airports dataset
airports_df1 <- airports_df[, c("iata", "state")]
airports_df2 <- airports_df[, c("iata", "state")]
```

We need 2 sets of the filtered airports dataset to get both origin and destination state

airports_df1 will be used to get the origin state

airports_df2 will be used to get the destination state

```{r}
# Renaming the columns in airports_df1
colnames(airports_df1) <- c("Origin", "Origin_State")
airports_df1
```

```{r}
# Renaming the columns in airports_df2
colnames(airports_df2) <- c("Dest", "Destination_State")
airports_df2
```

```{r}
# Merging airports_df1 and the main dataset to get the origin for each flight
merged_df <- merge(cleaned_df, airports_df1, by = "Origin")
merged_df
```

```{r}
#Merging airports_df2 and the merged dataset to get the destination for each fligh
merged_df <- merge(merged_df, airports_df2, by = "Dest")
merged_df
```

```{r}
# Creating a new column "Route" that combines the Origin_State and Destination_State
merged_df$Route <- paste(merged_df$Origin_State, "-", merged_df$Destination_State)
merged_df
```

```{r}
# Filtering to keep only the required columns
merged_df <- merged_df[, c("Year", "Route")]
merged_df
```

```{r}
# Getting the count of each route and stroing it in a new dataframe in descedning orde rof count
count_of_routes <- as.data.frame(table(merged_df$Route)) %>%
  arrange(desc(Freq))
colnames(count_of_routes) <- c("Route", "Count")
count_of_routes
```

```{r}
# Creating a list with only the top 20 most used routes
top20_routes <- head(count_of_routes$Route, 20)
top20_routes
```

```{r}
# Keeping entries in the merged dataset with the top 20 most used routes
merged_df <- merged_df[merged_df$Route %in% top20_routes, ]
merged_df
```

```{r}
# Grouping the data by year and route and getting the count of how many flights were taken for a  sepcific route for each year
route_counts <- merged_df %>%
  group_by(Year, Route) %>%
  summarize(Count = n())
route_counts
```

```{r}
# Splitting the "route_counts" by years
y2006_df <- route_counts[route_counts$Year == 2006, ]
y2007_df <- route_counts[route_counts$Year == 2007, ]
```

```{r}
# Setting the positions on the x-axis for the bars of each year
x_positions_1 <- 1:length(y2006_df$Route)
x_positions_2 <- x_positions_1 + 0.4

# Creating a new plot
plot <- ggplot()

# Plotting the data for 2006
plot <- plot + geom_bar(data = y2006_df, aes(x = x_positions_1 - 0.2, y = Count),
                        width = 0.4, fill = "blue", alpha = 0.9, stat = "identity",
                        position = "identity", show.legend = TRUE)

# Plotting the data for 2007
plot <- plot + geom_bar(data = y2007_df, aes(x = x_positions_2 - 0.2, y = Count),
                        width = 0.4, fill = "red", alpha = 0.9, stat = "identity",
                        position = "identity", show.legend = TRUE)

# Setting the x-axis tick positions and labels
plot <- plot + scale_x_continuous(breaks = x_positions_1, labels = y2006_df$Route, 
                                   expand = c(0,0)) + labs(x = "Route")

# Changing the y-axis labels to regular numbers, if not they will appear as scientific like 1e+05 and so on
plot <- plot + scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE),
                                  limits = c(0, max(route_counts$Count) + 50000), breaks = seq(0, max(route_counts$Count) + 50000, 50000),
                                  expand = c(0, 0.1)) + labs(y = "Total Number of Flights")

# Adding the title for the plot
plot <- plot + labs(title = "Total Number of Flights for the 20 most flown routes")

# Flipping the x-axis labels so that to increase clarity
plot <- plot + theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Moving the title of the plot to the middle
plot <- plot + theme(plot.title = element_text(hjust = 0.5))

# Printingthe plot
print(plot)

# Saving the plot locally as a png
ggsave("Total Number of Flights for the 20 most flown routes - R.png")
```
