---
title: "Abdussamad - 190519205 - Question 2 - R"
author: "Abdussamad Roshan"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 2 - Do older planes suffer more delays?

## Importing and Loading Packages

```{r}
# # Installing the necessary packages
install.packages("dplyr")
install.packages("ggplot2")
install.packages("plotly")
install.packages("rmarkdown")
install.packages("tidyverse")
```

```{r}
# Loading the packages
library(dplyr)
library(ggplot2)
library(plotly)
library(rmarkdown)
library(tidyr)
library(tidyverse)
```

## Importing the Datasets

```{r}
# Importing the cleaned dataset
cleaned_df <- read.csv("Cleaned_Dataset_R.csv")
# Importing the Plane Data, any NA or empty values are replaced with NA
plane_df <- read.csv("plane-data.csv", na.strings = c("", "NA"))
```

```{r}
# Checking the first 5 rows of the cleaned dataset
head(cleaned_df, n = 5)
```

```{r}
# Showing first 5 rows of the plane dataset
head(plane_df, n = 5)
```

```{r}
#Filtering to keep only the neccessary columns in the main dataset
cleaned_df <- cleaned_df[, c("TailNum", "ArrDelay", "DepDelay")]
cleaned_df
```

```{r}
# Filtering to keep only the neccessary columns in the plane dataset
plane_df_filtered <- plane_df[, c("tailnum", "year")]
plane_df_filtered
```

```{r}
# Checking for null values in the filtered plane dataset
colSums(is.na(plane_df_filtered))
```

```{r}
# Dropping all rows with null values
plane_df_filtered <- na.omit(plane_df_filtered)
plane_df_filtered
```

```{r}
# Checking if all null values have been removed
colSums(is.na(plane_df_filtered))
```

```{r}
# Showing general information about the filtered plane data to check if data has been cleaned properly
str(plane_df_filtered)
```

We can see that the "year" columns' data type is chr which suggests uncleaned data as the data type should be integer

```{r}
# Getting the value counts for the "year" column
table(plane_df_filtered$year)
```

We can see that there are values "None" and "0000" which needs to be removed

```{r}
# Dropping "None" and "0000" from the "year" column
plane_df_filtered <- subset(plane_df_filtered, year != "0000" & year != "None")
plane_df_filtered
```

```{r}
# Checking if None" and "0000" has been removed
table(plane_df_filtered$year)
```

```{r}
# Converting the data type of the "year" column to integer
plane_df_filtered$year <- as.integer(plane_df_filtered$year)
```

```{r}
# Checking if the data type has been converted
str(plane_df_filtered)
```

The common column we want to merge on is the TailNum, but the column name differs between the two dataframes

```{r}
# Changing the "tailnum" column in the filtered plane dataframe to "TailNum" to make merging possible
names(plane_df_filtered)[names(plane_df_filtered) == "tailnum"] <- "TailNum"
plane_df_filtered
```

```{r}
# Merging the two dataframes on the"TailNum" column
merged_df <- merge(cleaned_df, plane_df_filtered, by = "TailNum")
merged_df
```

```{r}
# Checking the information of the merged datframe
str(merged_df)
```

```{r}
# Checking for null values in the merged dataset
colSums(is.na(merged_df))
```

```{r}
# Creating a new varibale to hold the mean average delay grouped by year
mean_ArrDelay_byyear <- merged_df[, c("year", "ArrDelay")] %>% 
  group_by(year) %>% 
  summarise(mean_ArrDelay = mean(ArrDelay))
mean_ArrDelay_byyear
```

```{r}
ggplot(data = mean_ArrDelay_byyear, aes(x = year, y = mean_ArrDelay)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Average delay for manufacture year of plane", 
       x = "Year of Manufacture", 
       y = "Average Delay in Minutes") + 
  theme(plot.title = element_text(hjust = 0.5))

# Saving the plot locally as a png
ggsave("Average Delay for Plane Year - R.png")
```
