---
title: "Abdussamad - 190519205 - Question 4 - R"
author: "Abdussamad Roshan"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 4 - Can you detect cascading failures as delays in one airport create delays in others?

## Importing and Loading Packages

```{r}
# Install the following packages if not already installed
install.packages("lubridate")
install.packages("stringi")
install.packages("Rtools")
install.packages("janitor")
install.packages("vtree")
install.packages('rmarkdown')
```

```{r}
# Loading the packages
library(dplyr)
library(ggplot2)
library(janitor)
library(lubridate)
library(plotly)
library(rmarkdown)
library(scales)
library(stringi)
library(tidyr)
library(tidyverse)
```

## Importing the Dataset

```{r}
# Importing the main cleaned dataset, any NA or empty values are replaced with NA
cleaned_df <- read.csv("Cleaned_Dataset_R.csv", na.strings = c("", "NA"))

cleaned_df
```

Using the "CRSDepTime", "DayofMonth", "Month" and "Year" columns, a new column is created to give the scheduled departure time of each flight in datetime format

```{r}
# Selecting the first 2 numbers gives the minute part of the CRSDepTime
CRSDepTime_Minutes <- stri_sub(as.character(cleaned_df$CRSDepTime), -2, -1)
```

```{r}
# Checking the type of CRSDepTime_Minutes
typeof(CRSDepTime_Minutes)
```

We need the type to be a list for future operations

```{r}
# Making a list with CRSDepTime_Minutes
List_CRSDepTime_Minutes <- as.list(CRSDepTime_Minutes)
```

```{r}
# Selecting the last 2 numbers gives the hour part of the CRSDepTime
CRSDepTime_Hours <- stri_sub(as.character(cleaned_df$CRSDepTime), 1, -3)
```

```{r}
# Checking the type of CRSDepTime_Hours
typeof(CRSDepTime_Hours)
```

```{r}
# Making a list with CRSDepTime_Hours
List_CRSDepTime_Hours <- as.list(CRSDepTime_Hours)
```

There will be entries where CRSDepTime_Hours will be empty, for example 30 which indicates 00:30 am, will not have an hour value. Therefore all such empty entries will be replaced with 0 to avoid errors.'

```{r}
# Replacing empty values with 0
List_CRSDepTime_Hours[List_CRSDepTime_Hours==""] <- 0
```

```{r}
# Getting the DayofMonth, converting it to a character and storing it in a list
List_Day <- as.list(as.character(cleaned_df$DayofMonth))
```

```{r}
# Getting the Month, converting it to a character and storing it in a list
List_Month <- as.list(as.character(cleaned_df$Month))
```

```{r}
# Getting the Year, converting it to a character and storing it in a list
List_Year <- as.list(as.character(cleaned_df$Year))
```

Adding a column to full_dataset which contains a string with information about the Date and Time of estimated departure of each flight

```{r}
# Creating the new column called "DateTimeCRSDepTime"
# This holds a string containing the Day, Month, Year, Hours and Minutes seperated by a "-"
cleaned_df["DateTimeCRSDepTime"] <- paste(List_Day, List_Month, List_Year, List_CRSDepTime_Hours, List_CRSDepTime_Minutes, sep="-")
```

```{r}
# Removing data not needed anymore to avoid RAM bottlenecks
rm(List_CRSDepTime_Hours)
rm(List_CRSDepTime_Minutes)
rm(List_Day)
rm(List_Month)
rm(List_Year)
rm(CRSDepTime_Hours)
rm(CRSDepTime_Minutes)
```

```{r}
# Converting "DateTimeCRSDepTime" to date-time
cleaned_df$DateTimeCRSDepTime <- as.POSIXct(cleaned_df$DateTimeCRSDepTime, format="%d-%m-%Y-%H-%M")
```

```{r}
# Examining the class of "DateTimeCRSDepTime" column
class(cleaned_df$DateTimeCRSDepTime)
# Removing class attribute to get the numeric value
unclass(cleaned_df$DateTimeCRSDepTime[[1]])
```

```{r}
# Creating new column "Total_Delay" that adds up the arrival delay and departure delay 
cleaned_df["TotalDelay"] <- cleaned_df$ArrDelay + cleaned_df$DepDelay
```

```{r}
# Sorting the flights by "DateTimeCRSDepTime"  and grouping by TailNum
cleaned_df <- cleaned_df %>% arrange(DateTimeCRSDepTime) %>% group_by(TailNum)
```

Shifting the values in each group down by one position, so that each row has the "TotalDelay" value of the previous row

```{r}
# Creating a new column called "PreviousDelay"
cleaned_df["PreviousDelay"] <- cleaned_df$TotalDelay
```

```{r}
# Moving "PreviousDelay" down by 1 row
cleaned_df <- cleaned_df %>% mutate_at(c("PreviousDelay"), list(lag), n = 1)
```

```{r}
# Checking for any null values
colSums(is.na(cleaned_df))
```

There are multiple reasons for the existence of null values 1. If a plane has had only 1 recorded flight it will not have a "PreviousDelay" 2. The first flight for any individual plane will also not have any "PreviousDelay"

```{r}
# Dropping null values and checking if they have been removed
cleaned_df <- na.omit(cleaned_df)
colSums(is.na(cleaned_df))
```

## Plotting a scatter plot with a regression line to show how the delay of the previous flight affects the delay of the next flight

```{r}
# Plotting a scatter plot with a linear regression line
ggplot(data = cleaned_df, aes(x = PreviousDelay, y = TotalDelay)) + 
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  
  # Setting the x-axis and y-axis labels and the title
  labs(x = "Delay of previous flight in minutes", y = "Delay of next flight in minutes",
       title = "Scatter plot for Previous Delay vs Current Delay") +
  
  theme(plot.title = element_text(size = 15, hjust = 0.5), 
        axis.title.x = element_text(size = 13), 
        axis.title.y = element_text(size = 13))

# Saving the plot locally as a PNG
ggsave("Previuos Delay vs Next Delay - R.png")
```

## Getting a Contingency table for "PresenceOfCurrentDelay" vs "PresenceOfCurrentDelay"

```{r}
# Creating a new column "PresenceOfCurrentDelay", if the "TotalDelay" is greater than 0 "Present" is entered and if not "Absent" 
cleaned_df$PresenceOfCurrentDelay <- ifelse(cleaned_df$TotalDelay > 0, "Present", "Absent")
```

```{r}
# Creating a new column "PresenceOfPreviousDelay", if the "PreviosDelay" is greater than 0 "Present" is entered and if not "Absent"
cleaned_df$PresenceOfPreviousDelay <- ifelse(cleaned_df$PreviousDelay > 0, "Present", "Absent")
```

```{r}
cleaned_df
```

```{r}
# Getting a contingency table and getting the percentages up to 2 decimal places

crosstab <- tabyl(cleaned_df, PresenceOfPreviousDelay, PresenceOfCurrentDelay) %>% adorn_percentages("row") %>% adorn_pct_formatting(digits = 2)
crosstab
```

# 
